An Experimental Study of Sorting and Branch Prediction

### 2007-01-18 docs/branch_paper/paper.tex
Moved nicks paper into the branch_paper directory

\author{
PAUL BIGGAR
\and
DAVID GREGG
\\University of Dublin, Trinity College
}

\begin{abstract}

\end{abstract}

@@

\begin{bottomstuff}
Corresponding author's address: David Gregg, Department of Computer
Science, University of Dublin, Trinity College, Dublin 2, Ireland. {\tt
David.Gregg@cd.tcd.ie}.
\end{bottomstuff}

@@

\section{Motivation}
Classical analyses of algorithms make simplifying assumptions about
the cost of different machine instructions. For example, Knuth's Mix
machine code assumes that all operations have unit costs. More
recently researchers have recognized that on modern computers the cost
of accessing memory can vary dramatically depending on whether the
data can be found in the first-level cache, or must be fetched from a
lower level of cache or even main memory.  This has spawned a great
deal of research on cache-efficient searching and sorting \cite{}.

Another type of instruction whose cost can vary dramatically is the
conditional branch. Modern pipelined processors depend on \emph{branch
prediction} for much of their performance.  If the direction of a
conditional branch is correctly predicted ahead of time, the cost of
the conditional branch may be as little as the cost of, say, an
integer add instruction. If, on the other hand, the branch is
mispredicted the the processor must flush its pipeline, and restart
from the correct target of the branch. This cost is typically a large
multiple of the cost of executing a correctly-predicted branch. For
example, the Intel Pentium 4 processor has a 31-stage pipeline,
meaning that a branch misprediction will cost around 30 cycles.  In
contrast, on the same processor the cost of a memory access that
misses the first level cache and hits the second level cache is 23
cycles. Fortunately, the branches in most programs are very
predictable, so branch mispredictions are usually rare (prediction
accuracies of greater than 90\% are typical).

The cost of executing branches is particularly important for sorting
because the inner loops of most sorting algorithms consist of
camparisons of items to be sorted. Thus, the predictability of
these comparison branches is critical to the performance of
sorting algorithms. In this paper we study the predictability
of branches is most of the major sorting algorithms. We also
examine cache-conscious variations of some of the algorithms,
and show how optimizations for the cache impact in branch
mispredictions.

The rest of this paper is organized as follows.

\section{Branch prediction}

\section{Experimental setup}

\section{Elementary sorts}
\label{ordernsquared}
Elementary sorts are also referred to as $O(N^2)$ sorts. These
algorithms have the worst performance for large inputs. However they
have two advantages: they are simple to implement correctly and they
are often faster than more sophisticated algorithms for small input
sets. Four of these sorts are discussed here: selection sort,
insertion sort, bubblesort and shakersort.

Selection sort is a simple algorithm that repeatedly finds the
smallest element of the array and removes it to a sorted section at
the start. It requires approximately $N^2/2$ comparisons and exactly
$N$ exchanges. Insertion sort takes a different strategy and instead
creates an (initially empty) sorted region at the start of the array
and inserts items into their correct position in this sorted region.
Insertion sort performs in linear time on a sorted list. With a random
list, there are, on average, there are $N^2/4$ comparisons and $N^2/4$
half-exchanges. Thus, selection sort performs around twice as many
branches as insertion sort.

Although we know the approximate number of comparisons that each
algorithm performs, and important question is whether there is any
difference of the predictability of the branches in each algorithm.
Figure \ref{} shows the number of mispredictions per key for both the
algorithms using random data, and varying data set sizes, measured
using the Simplescalar processor simulator.

There are three notable features in this data. Firstly, the number of
mispredictions is very low. If the behaviour of these branches were
random, then we would expect that half of the comparison branches
would be mispredictions, that we would expect $O(N^2)$
mispredictions. In fact, the number of mispredictions per key is less
than 12 even for the largest arrays. Secondly, although selection sort
performs twice as many comparison branches it does not cause twice as
many branch mispredictions. For the bimodal predictor, it causes 9-12
times as many mispredictions within the size of input sets that we
examine. Thirdly the two-level branch predictor is slightly worse at
predicting the comparison branches in these algorithms than the simple
bimodal predictor.  This is a remarkable result, because two-level
predictors are similar to, or sigificantly more accurate than bimodal
predictors for almost all other types of branches \cite{Uht}.

\subsection{Insertion sort}
With a bimodal predictor, insertion sort causes almost exactly one
branch misprediction per key, regardless of the size of the array.
Thus, although insertion sort performs around $N/4$ comparisons per
key, only one of these comparisons causes a misprediction. The
reason for this becomes clear when we consider the inner loop
of insertion sort:
\begin{verbatim}
   while( item < a[j-1] ) {
     a[j] = a[j-1];
     j--;
   }
   a[j] = item;
\end{verbatim}

The comparison in insertion sort is also the loop guard.  As long as
the loop continues, this comparison always resolves in the same
direction, and is thus highly predictable. When the location for the
item is finally found, the comparison resolves in the opposite
direction, and a branch misprediction occurs. Thus, for a bimodal
predictor, the number of mispredictions per key is almost exactly
one. Although the results we present are for random data, the same
is true regardless of the input data. Insertion sort will always
cause around one misprediction per key.

The two-level predictor combines a pattern of recent branch outcomes
make a prediction. This goal is to exploit correlations between branch
outcomes. For most programs this is valuable, but there is correlation
of branch outcomes in insertion sort. The additional complexity simply
causes the predictor to find false correlations, with the result that
number of mispredictions is higher than for a simpler, bimodal
predictor. In general, misprediction rates of more than one per key
are caused by imperfections in the predictor, rather than the
unpredictability of the comparison branch in insertion sort.

\subsection{Selection sort}
The comparison branch in selection sort is also highly predictable,
but less so than for insertion sort. The inner loop of selection
sort consists of searching for the smallest element of part of
the array:
\begin{verbatim}
  min = i;
  for ( j = i+1; j < N; j++ )
    if ( a[j] < a[min] ) min = j;
\end{verbatim}
The predictability of the comparison branch depends on the number of
times that the cadidate minimum, $min$, changes. Knuth \cite{Knuth97}
shows that for $N$ elements, the number of changes is $H_N - 1$, where
$H_N$ is the harmonic series:
$$H_N = 1 + \frac{1}{2} + \frac{1}{3} + ... + \frac{1}{N} =  \sum_{k=1}^{N}
\frac{1}{k},  \hspace{0.4cm}  N \geq 1$$

Within practical sorting ranges, $H_N$ grows at a rate similar to
$log_e(N)$. Thus, in the average case, we can expect that the comparison
branch will mostly go in the same direction, making the branch highly
predictable.

The number of changes to the left-to-right minimum in the best case is
zero and in the worst case $N-1$. In the worst case, the array is
sorted in reverse order, and so the left-to-right minimum changes on
every element. Interestingly, the worst cases for the number of
changes is not the worst case for branch mispredictions. If the
minimum changes for every element, then the comparison branch will
always resolve in the same direction, and thus be almost perfectly
predictable. From the point of view of branch prediction, frequent
changes (preferably with no simple pattern) are the main indicator of
poor performance, rather than the absolute number of times that the
branch resolves in each direction. The worst case for selection sort
would be an array which is partially sorted in reverse order; the
left-to-right minimum would change frequently, but not predictably so.

\subsection{Bubble sort}
Bubblesort is a notoriously inefficient algorithm that compares and
exchanges adjacent pairs of elements in the array. One execution of
the inner loop will move the smallest element of the array to the
first position, and partially sort the rest of the array by moving
small elements towards the start. Cocktail shaker sort (or simply
shakersort) alternates the direction of the inner loop to take
more advantage of this partial sorting to reduce the number of
outer-loop iterations.

Figure \ref{} shows the number of branch mispredictions per key for
bubble- and shakersort.  Nearly one in four comparisons in bubblesort
are mispredicted. As the data sets gets large, there are nearly 10000
branch mispredictions per key key. Shakersort executes a lot fewer
branches, and so has fewer mispredictions, but the rate of
misprediction is similar.

The branch behaviour of bubblesort is particularly interesting when
compared to selection sort. Although we usually think of the two
algorithms as being different, they are actually very similar. Like
selection sort, bubblesort finds the smallest key in the
unsorted part of the array and moves it to start of the
array. Thus, we would expect that bubblesort's branch prediction
behaviour would be similar to that of selection sort.

However, in the process of moving the smallest element into position,
bubblesort also rearranges other array elements. Selection sort sweeps
across the array and always keeps track of the left-to-right minimum
up to the current point.  However, bubblesort actually moves this
minimum, exchanging pairwise with each element it encounters which is
larger. Thus, even the unsorted part of the array becomes partially
sorted. We might expect that bubblesort's comparison branch for a
partially sorted array might be more predictable than for a random
one, but it is not. With a random array, the smaller elements are
equally likely to appear anywhere in the array. With a partially
sorted array, the smaller elements tend to appear towards the start of
the array which is, in bubblesort, the last place that we look. Thus,
the candidate minimum is changed much more often than if we were
searching a randomly ordered array.

\begin{figure}
\includegraphics{plots/two_point_counter_bubblesort.eps}
\caption{Behaviour of branches in bubblesort compared to the ``sweep'' number}
\label{Predictability-of-branches-in-bubblesort-compared-to-sweep-number}
\end{figure}

To demonstrate this effect, we measured how the predictability of
bubblesort's comparison branch varies with successive sweeps across
the array. We added a simple software simulation of a bimodal
predictor to the source code of base bubblesort, and ran it on a large
number of randomly generated arrays, each with 8192 elements.

Figure
\ref{Predictability-of-branches-in-bubblesort-compared-to-sweep-number}
shows that on the early iterations the number of correct predictions
is very close to the number of executed comparison branches. The
comparison branch is highly predictable because bubblesort's search
for the smallest element is similar to selection sort. However, with
successive sweeps even the unsorted part of the array becomes
increasingly sorted and the branch becomes more unpredictable. We see
that the number of executed comparison branches falls linearly, but
that the number of correctly predicted comparison branches falls much
more quickly. Eventually, the array becomes so close to fully sorted
that the branch becomes increasingly predictable in the other
direction, and the number of correct predictions again approaches the
number of executed branches. However, during the intermediate stage,
a huge number of branch mispredictions have already
occurred. Paradoxically, by partially sorting the array, bubblesort
greatly increases the number of branch mispredictions and thus greatly
increases its running time.

Overall, our results show that branch prediction is an important
factor in the efficiency of sorting algorithms. For example,
shakersort executes fewer machine instructions and causes fewer cache
misses than selection sort, but is slower due to branch
mispredictions. Insertion sort is confirmed as the most efficient
elementary sorting algorithm. In additon to having the best time
complexity it causes only around one branch misprediction per key even
in the worst case.

\section{Shellsort}
Shellsort \cite{Shell59} was the first in-place sub-$N^2$ sorting
algorithm. The algorithm operates by sorting all elements that are $h$
elements apart using insertion, where $h$ is some increment.  On early
iterations the increment is large, allowing items to move large
distances. The increment is reduced on successive iterations to allow
more local sorting. The final iteration is a standard insertion sort,
after which the array is sorted. We use Gonnet's increments, which set
the next value of the increment to $\frac{5}{11}^{th}s$ of the
previous increment, because we found them to be efficient in practice.
All discussions of complexity are based on these increments.

The branch performance of shellsort should be predictable. There are
$log_{(\frac{11}{5})}N$ increments, each with a separate pass over the
array. On each pass, all keys are insertion sorted into some
sub-array. Given that insertion sort causes around one branch
misprediction per key, we can expect $N$ comparison branch
mispredictions per pass. It is expected, therefore, that there will be
around $log_{(\frac{11}{5})}N$ mispredictions per key. In general, if
there are $i$ increments, shellsort will cause $ni$ mispredictions.

Figure \ref{} shows the number of branches and branch mispredictions
per key for shellsort. For 4194304 keys, we would expect
$log_{(\frac{11}{5})}4194304 = 18.34$ mispredictions per key. The
number of mispredictions we measure in our simulation is actually a
little under 18, which is remarkably close to the predicted number.

To further investigate this we separated the comparison branches from
the other branches in the code in a separate simulation of a bimodal
branch predictor on a large number of randomly generated arrays. We
found that using Gonnet's increments each key moves only an average of
0.9744 places in each insertion sort (although moving one place
actually involves skipping over $h$ other array elements, where $h$ is
the increment). Thus, the situation is somewhat different to insertion
sorting large arrays of random elements where each element can be
expected to move a large number of places. Instead of the comparison
branch almost always resolving in one direction, the actual outcome
will be close to random. We can still expect a number of
mispredictions that is close to one per key, but we are unlikely to
get the almost perfectly predictable behaviour that we see when using
insertion sort where the average number of places moved is larger.

The main apparent weakness of shellsort on modern architectures is its
poor locality of reference in memory access.  By comparing items which
are separated by $h$ keys, there is little reuse of items in the same
cache line. One can easily imagine cases where shellsort would cause
very large numbers of cache misses.  In the worst case, each key would
move a very large number of places in each insertion sort pass,
possibly causing a cache miss for every place that it moves.  To
further investigate this effect, we measured the number of first and
second-level cache misses caused by our implementation of shellsort
(see figure \ref{}). In fact, the number of mispredictions caused by
our shellsort is significant, but not much greater than other sorting
algorithms such as heapsort and merge sort (see sections \ref{} and
\ref{}). The main reason is that using Gonnet's increments on average
keys move only 0.9744 places on each pass, so very large numbers of
references in diverse parts of the array are not needed.

\section{Heapsort}
Heapsort is an in-place sort which works in $O(NlogN)$ time in the
average and worst case. It begins by creating a structure known as a
\n{heap}, a binary tree in which a parent is always larger its child,
and the largest key is at the top of the heap. It then repeatedly
removes the largest key from the top and places it in its final
position in the array, fixing the heap as it goes.

\section{Quicksort}
Quicksort is a divide and conquer algorithm that recursively
partitions its array around a pivot, swapping large keys on the left
hand side with smaller keys on the right hand side. The partition
causes the pivot to be put in its final position, with all greater
keys to the right, and lesser keys on the left. The left and right
partitions are then sorted recursively.

\subsection{Simple Quicksort}
Sedgewick \cite{} has proposed a number of improvements to the
standard quicksort algorithm, which we use in our base quicksort
implementation. First, we use an explicit stack to save space.
Secondly, we always sort the smaller partition of the array first, to
reduce the worst case size of the stack (from $O(N)$ to
$O(logN)$). Thirdly, we not quicksort patitions of less than a certain
\texttt{THRESHOLD} size; instead we do a final insertion pass over the
array after quicksorting. The smallest of the first \texttt{THRESHOLD}
keys is placed in the first location of the array as a
sentinel. Finally, we use median-of-3 to select the pivot, which
greatly reduces the likelihood of very unbalanced partitioning. This
also allows the bounds check to be removed from the inner loop,
greatly reducing the instruction count. Improvements such as these are
found in most efficient implementations of quicksort.

In addition to our base quicksort algorithm, we experimented with
variations which attempt to reduce the number of data cache
misses. LaMarca \cite{} proposed a \textit{memory-tuned} quicksort.
Once a partition is reduced below the size of the threshold, insertion
sort is preformed immediately. By doing this, the keys being sorted
are in the cache already. By leaving the insertion sort until the end,
as Sedgewick does, the reduction in instruction count is made up by
the increase in cache misses, once the lists being sorted are larger
than the cache.

\subsection{Multi-Quicksort}
Generally, quicksort has good locality of reference. It divides the
same region of the array into increasingly small partitions.  In the
case that the array to be sorted fits in a particular level of the
cache, then quicksort has excellent cache properties at that
level. Once the array is larger than the cache, however, the number of
misses increases dramatically. On the early passes, which divide the
array into large sections, there is likely to be little reuse of data
cache entries between passes. Given that there is little reuse between
passes, it would be better to reduce the number of passes by dividing
the array into a larger number of partitions on each pass. LaMarca
uses a technique similar to that used in multi-mergesort, in that a
\n{multi-way partition} is done at the start, moving all keys into
cache sized containers which can be sorted individually.

Implementing a multi-way partition is complicated. At the start of the
algorithm, a set of pivots is chosen and sorted, and keys are put into
containers based on which two pivots their value lies between. A
minimal and maximal key are put on the edges of this list to avoid
bounds checks as the list is iterated over.

The main problem with multi-quicksort is that we don't know how many
keys there will be in each container. We implement the containers as a
linked list of arrays. LaMarca notes very little performance
difference between 100 and 5000 keys in each block. The size is
therefore chosen so that each list fits exactly into a page of
memory. When a list is filled, another list is linked to it, and these
lists hold all the keys in the container. The total number of
containers is chosen so that on average each will hold $C/3$ keys,
where $C$ is the number of keys which fit in the cache. LaMarca shows
that using this formula on average only 5\% of subsets will be larger
than the cache.

The entire array is iterated though a key at a time. For each key, the
list of pivots are searched until the appropriate container is found,
and the key is added to that container. Sequentially searching the
lists increases the instruction count by 50\% over the memory-tuned
version. For this reason, the search was replaced with an efficient
binary search, which reduced by half the extra instructions.

Once the entire array has been put into containers, each container is
emptied in turn and the keys put back into the array. When a container
is empty, the pivot greater than that container is put into position,
which is guaranteed to be its final position. The indices of these
positions is pushed onto the stack, and the sub-array is then sorted.

While emptying the containers back into the array, an opportunity is
taken to find the smallest key in the leftmost partition. This is
placed as a sentinel, ensuring that every partition has a sentinel,
and that no bounds checks are required.

\subsection{Quicksort Results}
Figure \ref{} shows the

\section{Radixsort}

\section{Related work}

>>>>

\author{
PAUL BIGGAR\footnote[1]{PUT IN IRCSET GRANT DETAILS}, NICHOLAS NASH\footnotemark[1] KEVIN WILLIAMS\footnotemark[1] and DAVID GREGG
\\University of Dublin, Trinity College
}

\begin{abstract}
Sorting is one of the most important and well studied problems in Computer Science. Many good algorithms
are known which offer various trade-offs in efficiency, simplicity, memory use, and other factors. However,
these algorithms do not take into account features of modern computer architectures that significantly
influence performance. Caches and branch predictors are two such features, and while
there has been a significant
amount of research into the cache performance of general purpose sorting algorithms, 
there has been little research on their branch prediction properties. In this paper we empirically examine
the behaviour of the branches in all the most common sorting algorithms. 
We also consider the interaction of cache optimization on the predictability of the branches in these algorithms.
We find insertion sort to have the fewest branch mispredictions of any
comparison-based sorting algorithm, that bubble and shaker sort operate in a fashion which makes their
branches highly unpredictable, that the unpredictability of shellsort's branches improves
its caching behaviour and that several cache optimizations have little effect on mergesort's branch
mispredictions. We find also that optimizations to quicksort -- for example the choice of pivot -- have a strong
influence on the predictability of its branches. We point out a simple way of removing branch instructions from 
a classic heapsort implementation, and show also that unrolling a loop in a cache
optimized heapsort implementation improves the predicitability of its branches. Finally, we note
that when sorting random data two-level adaptive branch predictors are usually no better than simpler bimodal predictors. This is despite 
the fact
that two-level adaptive predictors are almost always superior to bimodal predictors in general.
\end{abstract}

@@

\begin{bottomstuff}
Corresponding author's address: David Gregg, Department of Computer
Science, University of Dublin, Trinity College, Dublin 2, Ireland. {\tt
David.Gregg@cs.tcd.ie}.
\end{bottomstuff}

@@

\section{Motivation}
Classical analyses of algorithms make simplifying assumptions about
the cost of different machine instructions. For example, the RAM model used for 
establishing asymptotic bounds and Knuth's MIX 
machine code both make drastically simplifying assumptions about the cost of
machine instructions \cite{KnuthVol1_97}. 
More
recently researchers have recognized that on modern computers the cost
of accessing memory can vary dramatically depending on whether the
data can be found in the first-level cache, or must be fetched from a
lower level of cache or even main memory.  This has spawned a great
deal of research on cache-efficient searching and sorting \cite{Nyberg+94,Agarwal96,LaMarca96b,LaMarca96a,LaMarca97,Xiao+00,Rahman+01,Wickremesinghe+02}.

Another type of instruction whose cost can vary dramatically is the
conditional branch. Modern pipelined processors depend on \emph{branch
prediction} for much of their performance.  If the direction of a
conditional branch is correctly predicted ahead of time, the cost of
the conditional branch may be as little as the cost of, say, an
integer add instruction. If, on the other hand, the branch is
mispredicted the processor must flush its pipeline, and restart
from the correct target of the branch. This cost is typically a large
multiple of the cost of executing a correctly-predicted branch. For
example, Intel Pentium 4 processors \cite{Intel248966-010,Intel249438-01} have pipelines of up 
to 31 stages,
meaning that a branch misprediction will cost around 30 cycles.  In
contrast, on the same processor the cost of a memory access that
misses the first-level cache and hits the second level cache is 23
cycles. Fortunately, the branches in most programs are very
predictable, so branch mispredictions are usually rare. Indeed, prediction
accuracies of greater than 90\% are typical \cite{Uht+97} with the
best predictors.

The cost of executing branches is particularly important for sorting
because the inner-loops of most sorting algorithms consist of
comparisons of items to be sorted. Thus, the predictability of
these comparison branches is critical to the performance of
sorting algorithms. In this paper we study the predictability
of branches in most of the major sorting algorithms. 
We focus on the behaviour of the branches whose outcome depends on a
comparison of keys presented to the sorting algorithm in its input. Throughout this paper we refer to
such branches as \textit{comparison branches}. Branches associated with controlling simpler aspects 
of the control flow of the algorithms are of much less interest because they are generally almost perfectly predictable.
We also examine cache conscious variations of many of the algorithms,
and show how optimizations for the cache influence the level of branch
mispredictions.

\section{Branch prediction}
\label{branch_prediction}

Branches are a type of instruction used for defining the flow control of programs. In high level languages they result from 
the use of statements like \texttt{if}, \texttt{while} and their variants. Branches can be either \textit{taken}, indicating that the
address they provide is the new value for the program counter, or they can be \textit{not-taken}, in which case sequential execution continues
as though the branch had not been present.

Branch instructions pose a difficulty for pipelined 
processors. In a pipelined processor, while the program counter points to a particular instruction, a potentially large
number of subsequent instructions are in a partially completed state.
Until it is known whether a branch is taken or not-taken, it is not possible to know what the next instructions to be executed are.
If the processor cannot tell what these next instructions are, it cannot fill its pipeline.
To keep the utilization of the pipeline at a reasonable level, modern processors attempt to anticipate the outcome of 
branch instructions by employing \textit{branch predictors}. A good discussion of pipelining and its associated issues can
be found in Hennessy and Patterson \citeyear{HennessyPatterson90}.

There are several kinds of branch predictors: static, semi-static and dynamic. We use only dynamic predictors in our experiments, 
because they are most commonly used in real processors. The statistics in the following paragraphs are taken from those derived
by Uht \textit{et al} \citeyear{Uht+97}, over the SPECint92 benchmarking suite.

The simplest type of dynamic predictor is referred to 
as a \textit{1-bit} predictor. It keeps a table recording, at each entry, whether a particular branch was taken or not-taken. 
Using the table, it predicts 
that a branch
will go the same way as it went on its previous execution. 1-bit predictors achieve 77\% to 79\% accuracy.

We refer to a 2-bit dynamic predictor as a \textit{bimodal} predictor. A bimodal predictor operates in the same manner as a 1-bit predictor but each table entry
can be thought of  as maintaining a counter from 0 to 3. The counter decrements on each
taken branch (with the exception of when the count is 0), and increments on each not-taken branch (with the exception of when the count is 3). With
counts equal to 0 or 1 the next branch is predicted as taken, with counts of 2 or 3 the branch is predicted not-taken. 
When the counter is 0, we say the predictor is in the \textit{strongly taken} state, since the branch must be not-taken twice before the not-taken
direction will be predicted again. When the counter is 1 we say the predictor is in the \textit{taken} state. Similarly counts of 2 and 3 are referred
to as \textit{not-taken} and \textit{strongly not-taken} respectively. Bimodal predictors achieve 78\% to 89\% accuracy.

Finally, some branch predictors attempt to exploit correlations in branch outcomes to improve accuracy. A \textit{two-level adaptive} predictor maintains
a \textit{branch history register}. This register records the outcomes of a number of previous branch instructions. For example, a register contents 
of \texttt{110010} indicates that the 6 previous 
branch outcomes were taken, taken, not-taken, not-taken, taken and not-taken.
For each branch, this register is used to index a table of bimodal predictors.
The program counter can also be included in the index, either concatenated or XORed with the indexing register. Two-level adpative
predictors are accurate about 93\% of the time.

In all the above, due to the size of the predictor table branches can collide. That is, two 
different branches can map to the same table location, often reducing accuracy.

\section{Experimental Setup}

We used truly random data provided by Haahr \citeyear{Random} in all our experiments. We used ten chunks of random data containing $2^{22}$ (4194304) keys.
Where a particular experiment used only part of the keys in a chunk, it used the left-most keys.
Our results are averaged over these chunks of data.

Processors do not generally provide access to the behaviour of their caches or branch predictors. However, a software processor simulator can be used to 
provide measurements 
of cache and branch prediction behaviour.
Our branch prediction and caching results were generated using the SimpleScalar PISA simulator version 3 \cite{SimpleScalarTutorialv4}. 
We used \texttt{sim-cache} and 
\texttt{sim-bpred} to generate results for caching and branch prediction characteristics respectively.
We used a variety of cache configurations; generally speaking we used an 8 KB level 1 data cache, 8 KB level 1 instruction cache,
and a shared 2 MB instruction and data level 2 cache, all with 32 byte cache lines. 
We also experimented with both direct mapped and fully associative caches. When presenting results
where the cache configuration is important, we will describe the precise configuration used. We present
results here only for direct mapped caches. The results for fully associative caches are similar \cite{BiggarGregg05}.

For the measurements taken using SimpleScalar, we used power of two sized sets of random keys ranging from $2^{12}$ to $2^{22}$ keys in size.
This is with the exception of the quadratic time sorting algorithms, for which the maximum set size used was $2^{16}$ keys. This was done
because of the very significant time it takes such algorithms to sort large inputs.

It takes time to fill arrays, and this can distort the relationship between the results for small set sizes, and larger set sizes, 
for which this time is amortised. As a result, we measured this time, and subtracted it from our results.
However, this practice also distorts the results. It removes compulsory cache misses from the results, so long as the data fits in the cache. 
When the data does not fit in the cache, the keys from the start of the array are faulted, with the effect that capacity misses reoccur, 
which are not discounted.

For branch prediction we simulated both bimodal and two-level adaptive predictors. 
We used power of two sized tables for the branch predictors. 
We used tables containing between $2^{11}$ and $2^{14}$ predictors.
When
presenting branch prediction results, the predictor configurations will be described precisely.
SimpleScalar's \texttt{sim-bpred} provides total results over
all the branches in a program. It is often desirable to examine particular branches individually however.
In order to do so, we added our own simulations of branch predictors to the programs. For our 
own simulations we average the results of each branch over ten runs of random data containing $2^{22}$ keys.

Finally we used \textit{PapiEx} \cite{PapiExManual} to access performance counters on a Pentium 4 1.6 Ghz. This allows cycle counts (i.e. running times)
to be
determined for the sorting algorithms on a real processor.
The results presented from these tests are averaged over 1024 runs and use the system's random number generator to provide data. The data
ranged in size from $2^{12}$ to $2^{22}$ keys in size.
The processor we used had an 8-way associative 256KB level 2 
cache with 64-byte cache lines. The separate data and instruction level 1 caches are 4-way associative, 8KB in size and have 32-byte cache lines.
The Pentium 4 uses an unspecified form of dynamic branch prediction. A reference manual \cite{Intel248966-010} mentions a branch history
register, so it is probable that the scheme is similar to two-level adaptive.

\section{Elementary sorts}

We begin by analysing the branch prediction behaviour of a number
of common elementary sorting algorithms, each operating in
$O(n^2)$ time. The main practical advantage of these
algorithms is their simplicity, which generally allows them to
operate faster than algorithms with $O(n \lg n)$ time complexity
for small inputs. Indeed, as we will see later, optimized versions
of $O(n \lg n)$ sorts use calls to elementary sorts for small inputs.
Knuth \citeyear{KnuthVol3_98} provides a good exposition of these elementary
sorting algorithms (as well as basic versions of the sorting algorithms
we shall examine in later sections).

\subsection{Selection Sort}
\label{selection_sort}

Selection sort is a very simple quadratic time sorting algorithm which begins by sequentially searching
for the smallest key in its input and moving it into position. The algorithm 
continues by repeatedly searching for and then moving the smallest key in the unsorted
section of its input into the appropriate position. The inner-loop of selection sort,
which is run a total of $n - 1$ times on an input \texttt{a[0..n - 1]} is shown below

\begin{verbatim}
  min = i;
  for(j = i + 1; j < n; j++)
    if(a[j] < a[min]) min = j;
  swap(a[i], a[min]);
\end{verbatim}

It is easy to see that selection sort performs approximately $n^2/2$ comparisons
and exactly $n - 1$ exchanges. 
On the $j^{th}$ iteration of the inner-loop the comparison branch is taken
if \texttt{a[j]} is the minimum of \texttt{a[i..j]}, which occurs with
probability $1 / (j - i + 1)$\footnote{Throughout this paper, for the purpose of analysis we make
the usual assumption that the input to be sorted consists of a random permutation of $\lbrace1, .., n\rbrace$.}. 
The expected number of times the branch is taken
on the first iteration is thus given by $H_n - 1$ where 
$H_n = \sum_{i = 1}^n 1/i$ is the $n^{th}$ harmonic number \cite{KnuthVol1_97}.
Since $H_n \approx \ln n$, and $\sum_{i = 1}^n H_i = O(n \ln n)$ the comparison branch
will be highly predictable. As expected, Figure \ref{elementary_sorts}(a) shows that number of mispredictions
per key grows in a similar fashion to $\ln n$.

If the array is initially in sorted order, then the value of \texttt{min}
will never change in the inner-loop. Since this implies the comparison branch
is never taken, it will be almost perfectly predictable. Similarly, 
if the array is sorted in reverse order, the value of \texttt{min} must
change at every element. That is, the comparison branch 
is taken at every element, and the branch is therefore almost perfectly
predictable.  From the point of view of branch prediction, frequent
changes (preferably with no simple pattern) are the main indicator of
poor performance, rather than the absolute number of times that the
branch resolves in each direction. The worst case for selection sort
would be an array which is partially sorted in reverse order; the value of
\texttt{min} would change frequently, but not predictably so.
Figure \ref{elementary_sorts}(a) shows the behaviour of selection
sort's branches.

\subsection{Insertion Sort}
\label{insertion_sort}

Insertion sort is another simple algorithm with quadratic
worst case time. It works by iterating the inner-loop below
a total of $n - 1$ times over its input

\begin{verbatim}
   item = a[i];
   while(item < a[i - 1]) 
   {
     a[i] = a[i - 1];
     i--;
   }
   a[i] = item;
\end{verbatim}
On the $i^{th}$ iteration \texttt{a[0..i - 1]} is sorted, allowing
\texttt{a[i]} to be placed in the correct position as shown in
the inner-loop above. On average, insertion sort performs approximately
$n^2/4$ comparisons and $n^2/4$ assignments in its inner-loop. Thus insertion sort
performs around half as many branches as selection sort.

A very pleasing property of insertion sort is that
it generally causes just 
a single branch misprediction per key, as Figure \ref{elementary_sorts}(a) shows. Typically
the inner \texttt{while} loop exit causes this misprediction, when
the correct location for the current item has been found.
In fact, a static
predictor would perform just as well as the two predictors shown, since it
would mispredict exactly once per key.

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.5\textwidth]{plots/selection_insertion_sort_branches.eps} & \includegraphics[width=0.5\textwidth]{plots/bubble_shaker_sort_branches.eps} \\
(a) & (b) \\
\end{tabular}
\caption{(a) Shows the branch mispredictions for selection and insertion sort using bimodal and two-level adaptive predictors both having 4096 table entries.
It is noteworthy the simple bimodal predictor has essentially identical performance to the two-level adaptive predictors.
(b) Shows the branch mispredictions for bubble and shaker sort using bimodal and two-level adaptive predictors, again both having 4096 table entries.
For bubble sort the bimodal and two-level adaptive predictors have close to identical performance. However, for shaker sort the simple bimodal
predictor results in fewer mispredictions than the two-level adaptive predictor. We also note that as a result of 
the partial sorting performed by bubble and shaker sort they incur far more
branch mispredictions per key than other $O(n^2)$ sorting algorithms.
\texttt{sim-bpred} was used to simulate the branch predictors in both (a) and (b). The two-level adaptive predictors used 10-bit history registers.}
\label{elementary_sorts}
\end{figure}

\subsection{Bubble sort}
\label{bubble_sort}

Bubble sort is another simple sorting algorithm, although notoriously
inefficient. It works by iterating the loop shown below at most $n - 1$
times on an input \texttt{a[0..n - 1]}.

\begin{verbatim}
for(j = 0; j < n - i - 1; j++)
{
    if(a[j + 1] < a[j])
       swap(a[j + 1], a[j]);
}
\end{verbatim} 
In this code $i$ is the outer-loop iteration counter, and begins at 0 and counts
towards $n - 1$. Bubble sort uses
information yielded by the comparisons which selection
sort simply discards. After a full iteration of the inner-loop shown larger
elements have moved to the right, closer to their final location.

Shaker sort is a variation of bubble sort in which there are two inner-loops
which are alternated. One is the inner-loop shown above. The other
inner-loop scans right-to-left moving small elements to the left in the
same manner as moving larger elements to the right has just been described.
Although shaker sort has the same worst-case time as bubble sort, it is generally
more efficient in practice.

Figure \ref{elementary_sorts}(b) shows the number of branch mispredictions per key for
bubble and shaker sort. As the data set gets large, bubble sort causes nearly 10000
branch mispredictions per key. Shaker sort executes a lot fewer
branches, and so has fewer mispredictions, but the rate of
misprediction is similar.

To understand the high misprediction rate of bubble sort we define $Q^k_j$
as the probability that $j^{th}$ inner-loop comparison branch is taken on the $k^{th}$ 
outer-loop iteration, $1 \leq k < n$. It turns out that

\begin{equation}
\label{bubble_equation}
Q^k_j = 
\left\{ 
\begin{array}{ll}
  \frac{j}{j + k} & \quad \mbox{if $j \leq n - k$}\\
  0 & \quad \mbox{otherwise}\\ 
\end{array} 
\right. 
\end{equation}

\noindent
This is a result of the fact that the $j^{th}$ comparison branch of the inner-loop
is taken only if \texttt{a[j]} is not larger than all of \texttt{a[0..j - 1]}.
For this
to be the case, the $(j + 1)^{th}$ branch in the previous outer-loop iteration must have been taken,
since otherwise the previous iteration of the outer-loop left the maximum of \texttt{a[0..j - 1]} in \texttt{a[j]}.
The probability that the $(j + 1)^{th}$ branch of the
previous outer-loop iteration is taken is $Q^{k - 1}_{j + 1}$. Given that the $(j + 1)^{th}$ branch of
the previous outer-loop iteration was taken, the probability that \texttt{a[j]} is
not larger than all of \texttt{a[0..j - 1]} is $1 - 1/(j + 1)$, thus,

\[
Q^k_j = Q^{k-1}_{j+1}\left(1 - \frac{1}{j + 1} \right)
\]

\noindent
The bases $Q^{1}_j = 1 - 1 / (j + 1)$ for $1 \leq j < n$ and $Q^1_n = 0$ give the solution shown in Equation \ref{bubble_equation}.
It is easy to see that in the first outer-loop iteration ($k = 1$) the comparison branch is very likely to be taken, whereas in the last
outer-loop iteration ($k = n - 1$), the comparison branch is very unlikely to be taken. Moving between these two extremes causes the intermediate
branches to be unpredictable. The remainder of this section examines this phenomenon in more detail.

In order to analyse the behaviour of bubble sort when a bimodal predictor is used,
it is useful to make use of the \textit{steady state predictability} function. This function
generally gives a good estimate of the probability that a particular branch will be correctly predicted,
given that we know the probability that the branch will be taken. For a bimodal predictor this function is
given by

\begin{equation}
CP(p) = \frac{3p^2 - 3p + 1}{2p^2 - 2p + 1}
\label{CP_equation}
\end{equation}
